---
title: "Bulk RNA-seq Differential Expression Analysis"
subtitle: "T. tenax Exposure in NCI-H292 Human Lung Epithelial Cells"
author: "Danya Hassan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
knitr::opts_knit$set(root.dir = normalizePath("."))
```

---

## 1. Overview

This notebook performs a complete bulk RNA-seq differential expression analysis
comparing human lung epithelial cells (NCI-H292) exposed to *Trichomonas tenax*
(*T. tenax*) protozoan against untreated controls (n = 3 per group).

**Upstream pipeline:** Raw FASTQ files were quality-trimmed, aligned to the
human reference genome (hg38) using HISAT2, and quantified with featureCounts
to produce the raw counts matrix used here.

**Analysis steps:**

1. Data loading and quality filtering  
2. DESeq2 normalization and DE analysis (with LFC shrinkage)  
3. QC visualizations (size factors, dispersion, PCA)  
4. Volcano plot and MA plot  
5. Gene ontology (GO Biological Process)  
6. KEGG pathway enrichment  
7. g:Profiler (GOST) multi-ontology enrichment  
8. Heatmap of top DE genes  
9. Pathway network visualization (cnetplot)

---

## 2. Libraries

```{r libraries}
# Bioconductor
library(DESeq2)
library(org.Hs.eg.db)
library(clusterProfiler)
library(EnhancedVolcano)
library(biomaRt)
library(ComplexHeatmap)

# CRAN
library(dplyr)
library(readr)
library(ggplot2)
library(ggrepel)
library(gprofiler2)
library(circlize)
library(tibble)
library(plotly)
library(RColorBrewer)
```

---

## 3. Load Data

```{r load-data}
# ── Edit paths here ──────────────────────────────────────────────────────────
COUNTS_PATH   <- "dh_counts_project.txt"
METADATA_PATH <- "dh_counts_project_metadata.csv"
# ─────────────────────────────────────────────────────────────────────────────

counts <- read.delim(
  COUNTS_PATH,
  comment.char    = "#",
  stringsAsFactors = FALSE,
  row.names        = 1
)

metadata <- read.csv(METADATA_PATH, stringsAsFactors = FALSE)

cat("Counts matrix dimensions:", nrow(counts), "genes ×", ncol(counts), "columns\n")
cat("Metadata rows           :", nrow(metadata), "\n")
```

---

## 4. Data Wrangling

### 4.1 Clean Counts Matrix

```{r clean-counts}
counts_clean <- counts %>%
  dplyr::select(-c(Chr, Start, End, Strand, Length)) %>%
  dplyr::rename(
    SRR23972383 = TT_Cultured_01.sam,
    SRR23972384 = TT_Cultured_02.sam,
    SRR23972385 = TT_Cultured_03.sam,
    SRR23972386 = Untreated_01.sam,
    SRR23972387 = Untreated_02.sam,
    SRR23972388 = Untreated_03.sam
  )

cat("Cleaned counts matrix:", nrow(counts_clean), "genes ×",
    ncol(counts_clean), "samples\n")
head(counts_clean, 3)
```

### 4.2 Format Sample Metadata

```{r clean-metadata}
sample_info <- metadata %>%
  dplyr::select(-geo_id) %>%
  dplyr::rename(ttenax = dex) %>%
  as.data.frame()

rownames(sample_info) <- sample_info$id
sample_info <- sample_info %>% dplyr::select(-id)

# Set factor levels explicitly: control as reference
sample_info$ttenax <- factor(sample_info$ttenax,
                             levels = c("control", "treated"))

sample_info
```

---

## 5. DESeq2 Analysis

### 5.1 Build DESeqDataSet and Filter

```{r deseq-build}
dds <- DESeqDataSetFromMatrix(
  countData = counts_clean,
  colData   = sample_info,
  design    = ~ ttenax
)

# Keep genes with > 10 total counts across all samples
dds_filtered <- dds[rowSums(counts(dds)) > 10, ]

cat("Genes before filtering:", nrow(dds), "\n")
cat("Genes after filtering :", nrow(dds_filtered), "\n")
```

### 5.2 Run DESeq2

```{r deseq-run}
dds_seq <- DESeq(dds_filtered)
```

### 5.3 Results with LFC Shrinkage

LFC shrinkage (apeglm method) reduces noise on low-count genes without
discarding them, producing more reliable fold-change estimates for ranking
and visualization.

```{r deseq-results}
# Raw results
res_raw <- results(dds_seq, alpha = 0.05,
                   contrast = c("ttenax", "treated", "control"))

# Shrunk LFC (recommended for visualization and ranking)
res_shrunk <- lfcShrink(dds_seq,
                         coef  = "ttenax_treated_vs_control",
                         type  = "apeglm",
                         res   = res_raw)

# Add gene symbols
res_df <- as.data.frame(res_shrunk)
res_df$symbol <- mapIds(
  org.Hs.eg.db,
  keys     = rownames(res_df),
  keytype  = "ENSEMBL",
  column   = "SYMBOL",
  multiVals = "first"
)

cat("Total tested genes    :", nrow(res_df), "\n")
cat("Significant (padj<0.05):", sum(res_df$padj < 0.05, na.rm = TRUE), "\n")
cat("Upregulated (LFC>1)   :",
    sum(res_df$padj < 0.05 & res_df$log2FoldChange > 1, na.rm = TRUE), "\n")
cat("Downregulated (LFC<-1):",
    sum(res_df$padj < 0.05 & res_df$log2FoldChange < -1, na.rm = TRUE), "\n")
```

---

## 6. QC Visualizations

### 6.1 Size Factor Check

Size factors close to 1.0 indicate similar sequencing depths across samples.
Large deviations (> 2×) may suggest normalization issues.

```{r size-factors}
sf_df <- data.frame(
  sample      = colnames(dds_seq),
  size_factor = sizeFactors(dds_seq),
  condition   = sample_info$ttenax
)

ggplot(sf_df, aes(x = sample, y = size_factor, fill = condition)) +
  geom_col(width = 0.6, color = "white") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "DESeq2 Size Factors per Sample",
       x = NULL, y = "Size Factor", fill = "Condition") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 6.2 Dispersion Estimates

A well-fitted dispersion trend (red curve) that follows the gene-wise estimates
(black dots) closely is expected. Outliers (circled) are genes with unusually
high variability.

```{r dispersion}
plotDispEsts(dds_seq,
             main = "Dispersion Estimates — T. tenax vs Control")
```

### 6.3 PCA — 2D (ggplot2)

```{r pca-2d}
vsd <- vst(dds_seq, blind = TRUE)

pca_data <- plotPCA(vsd, intgroup = "ttenax", returnData = TRUE)
var_exp  <- round(100 * attr(pca_data, "percentVar"), 1)

ggplot(pca_data, aes(x = PC1, y = PC2, color = ttenax, label = name)) +
  geom_point(size = 5, alpha = 0.9) +
  geom_text_repel(size = 3.5, show.legend = FALSE) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCA of VST-Normalized Counts",
    x     = paste0("PC1 (", var_exp[1], "% variance)"),
    y     = paste0("PC2 (", var_exp[2], "% variance)"),
    color = "Condition"
  ) +
  theme_minimal(base_size = 13)
```

### 6.4 PCA — 3D (Plotly)

```{r pca-3d}
counts_labeled <- counts_clean %>%
  dplyr::rename(
    "T-Tenax Treated 1" = SRR23972383,
    "T-Tenax Treated 2" = SRR23972384,
    "T-Tenax Treated 3" = SRR23972385,
    "Untreated 1"        = SRR23972386,
    "Untreated 2"        = SRR23972387,
    "Untreated 3"        = SRR23972388
  )

pca_3d     <- prcomp(t(counts_labeled), retx = TRUE)
pca_3d_df  <- as.data.frame(pca_3d$x[, 1:3])
pca_3d_df$Sample    <- rownames(pca_3d_df)
pca_3d_df$Condition <- ifelse(
  grepl("T-Tenax", pca_3d_df$Sample), "Treated", "Untreated"
)

plot_ly(
  pca_3d_df,
  x = ~PC1, y = ~PC2, z = ~PC3,
  type   = "scatter3d",
  mode   = "markers",
  color  = ~Condition,
  colors = c("Treated" = "steelblue", "Untreated" = "seagreen"),
  text   = ~Sample,
  marker = list(size = 7)
) %>%
  layout(title = "3D PCA — T. tenax Treated vs Untreated")
```

---

## 7. Differential Expression Plots

### 7.1 Volcano Plot

```{r volcano}
EnhancedVolcano(
  res_df,
  lab        = res_df$symbol,
  x          = "log2FoldChange",
  y          = "pvalue",
  title      = "T. tenax Treated vs Untreated — NCI-H292 Cells",
  subtitle   = "Shrunk LFC | padj < 0.05 | |LFC| > 1",
  pCutoff    = 0.05,
  FCcutoff   = 1.0,
  pointSize  = 2.5,
  labSize    = 3.5,
  colAlpha   = 0.6,
  legendPosition = "right",
  drawConnectors = TRUE,
  widthConnectors = 0.4
)
```

### 7.2 MA Plot

The MA plot displays log-fold change (M) against mean expression (A). After
LFC shrinkage, low-count genes (open circles) have their extreme folds pulled
toward zero, reducing false positives.

```{r ma-plot}
plotMA(res_shrunk,
       main  = "MA Plot — Shrunk LFC (apeglm)",
       ylim  = c(-5, 5),
       alpha = 0.05,
       colSig = "steelblue",
       colNonSig = "gray70")
abline(h = c(-1, 1), lty = 2, col = "crimson", lwd = 1.2)
```

---

## 8. Functional Enrichment

```{r sig-genes}
sig_df <- res_df %>%
  filter(padj < 0.05, !is.na(padj))

upreg_df <- sig_df %>% filter(log2FoldChange >  1.0)
dwreg_df <- sig_df %>% filter(log2FoldChange < -1.0)

cat("Significant upregulated genes  :", nrow(upreg_df), "\n")
cat("Significant downregulated genes:", nrow(dwreg_df), "\n")
```

### 8.1 GO Biological Process — Upregulated Genes

```{r go-upreg}
upreg_clean <- upreg_df %>% filter(baseMean > 100)

go_upreg <- enrichGO(
  gene          = rownames(upreg_clean),
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  readable      = TRUE
)

barplot(go_upreg, showCategory = 12) +
  ggtitle("GO Biological Process — Upregulated Genes") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 12))
```

### 8.2 GO Biological Process — Downregulated Genes

```{r go-dwreg}
dwreg_clean <- dwreg_df %>% filter(baseMean > 100)

go_dwreg <- enrichGO(
  gene          = rownames(dwreg_clean),
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  readable      = TRUE
)

if (!is.null(go_dwreg) && nrow(go_dwreg) > 0) {
  barplot(go_dwreg, showCategory = 12) +
    ggtitle("GO Biological Process — Downregulated Genes") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 12))
} else {
  cat("No significant GO BP terms found for downregulated genes.\n")
}
```

### 8.3 KEGG Pathway Enrichment — Upregulated Genes

```{r kegg-upreg}
upreg_entrez <- bitr(
  rownames(upreg_clean),
  fromType = "ENSEMBL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)

kegg_upreg <- enrichKEGG(
  gene         = upreg_entrez$ENTREZID,
  organism     = "hsa",
  keyType      = "ncbi-geneid",
  pvalueCutoff = 0.05
)

dotplot(kegg_upreg, showCategory = 15) +
  ggtitle("KEGG Pathway Enrichment — Upregulated Genes") +
  theme(plot.title = element_text(hjust = 0.5, size = 12))
```

### 8.4 Pathway Network Plot (cnetplot)

The cnetplot links enriched pathways (larger nodes) to their constituent genes
(smaller nodes), revealing shared genes across pathways.

```{r cnetplot, fig.width=10, fig.height=8}
# Convert ENSEMBL to symbol for readable labels
gene_lfc <- upreg_clean$log2FoldChange
names(gene_lfc) <- upreg_entrez$ENTREZID[
  match(rownames(upreg_clean), upreg_entrez$ENSEMBL)
]
gene_lfc <- na.omit(gene_lfc)

cnetplot(
  kegg_upreg,
  foldChange   = gene_lfc,
  showCategory = 6,
  circular     = FALSE,
  colorEdge    = TRUE
) +
  ggtitle("KEGG Pathway Network — Upregulated Genes") +
  scale_color_gradient2(
    low  = "steelblue",
    mid  = "white",
    high = "firebrick",
    midpoint = 0,
    name = "LFC"
  )
```

### 8.5 g:Profiler (GOST) — Upregulated Genes

```{r gost-upreg}
upreg_gost_genes <- sig_df %>%
  filter(log2FoldChange > 1.5, baseMean > 100) %>%
  rownames()

gost_upreg <- gost(upreg_gost_genes, organism = "hsapiens", significant = TRUE)

gostplot(gost_upreg, capped = TRUE, interactive = TRUE)
```

### 8.6 g:Profiler (GOST) — Downregulated Genes

```{r gost-dwreg}
dwreg_gost_genes <- sig_df %>%
  filter(log2FoldChange < -1.0, baseMean > 100) %>%
  rownames()

if (length(dwreg_gost_genes) > 0) {
  gost_dwreg <- gost(dwreg_gost_genes, organism = "hsapiens", significant = TRUE)
  gostplot(gost_dwreg, interactive = TRUE)
} else {
  cat("No downregulated genes met the filter criteria for GOST analysis.\n")
}
```

---

## 9. Heatmap of Top Differentially Expressed Genes

Z-score normalized expression across all six samples. Rows are clustered by
expression pattern; columns are clustered by sample similarity.

```{r heatmap, fig.width=9, fig.height=8}
heatmap_genes <- sig_df %>%
  filter(baseMean > 220, abs(log2FoldChange) > 1.3)

heatmap_counts <- counts(dds_seq, normalized = TRUE)[rownames(heatmap_genes), ]

# Z-score per gene
heatmap_z <- t(apply(heatmap_counts, 1, scale))
colnames(heatmap_z) <- c(
  "T-Tenax 1", "T-Tenax 2", "T-Tenax 3",
  "Untreated 1", "Untreated 2", "Untreated 3"
)

# Condition annotation bar
col_anno <- HeatmapAnnotation(
  Condition = c(rep("Treated", 3), rep("Untreated", 3)),
  col = list(Condition = c("Treated" = "steelblue", "Untreated" = "seagreen")),
  annotation_name_side = "left"
)

colors_heatmap <- colorRamp2(
  breaks = c(-2, 0, 2),
  colors = c("steelblue", "white", "firebrick")
)

Heatmap(
  heatmap_z,
  name              = "z-score",
  col               = colors_heatmap,
  top_annotation    = col_anno,
  cluster_rows      = TRUE,
  cluster_columns   = TRUE,
  show_row_dend     = TRUE,
  show_column_dend  = TRUE,
  row_labels        = heatmap_genes[rownames(heatmap_z), ]$symbol,
  row_names_gp      = gpar(fontsize = 7),
  column_names_gp   = gpar(fontsize = 10),
  border            = TRUE,
  column_title      = "Top DE Genes — T. tenax Treated vs Untreated (Z-score)",
  column_title_gp   = gpar(fontsize = 12, fontface = "bold")
)
```

---

## 10. Export Results

```{r export}
# Full DE results table
write.csv(
  res_df %>% arrange(padj),
  file      = "DEG_results_all.csv",
  row.names = TRUE
)

# Significant genes only
write.csv(
  sig_df %>% arrange(padj),
  file      = "DEG_results_significant.csv",
  row.names = TRUE
)

cat("Exported:\n")
cat("  DEG_results_all.csv         —", nrow(res_df), "genes\n")
cat("  DEG_results_significant.csv —", nrow(sig_df), "genes\n")
```

---

## 11. Session Info

```{r session-info}
sessionInfo()
```
